---
title: Hello Tidyverse and Tidymodels
author: Gabe Mednick
date: '2020-09-03'
slug: hello-tidyverse
categories: []
tags: []
subtitle: 'Work in progress'
summary: 'Data wrangling and visualization: insights into data analysis with the tidyverse'
authors: []
lastmod: '2020-10-20'
featured: no
disable_jquery: no
image:
  caption: '[Photo by Jesse Gardner on Unsplash](https://unsplash.com/photos/JYU14tnbfOA)'
  focal_point: ''
  preview_only: no
projects: []
---
# Intro 
This post uses data from the <i class='fab fa-r-project'  style='color: #367588'></i> for Data Science online learning community. Each Tuesday, a dataset is posted and people share their analyses and visualizations on Twitter with the hashtag, #TidyTuesday. This week's dataset (September 1st, 2020) is from [Our World in Data] (https://ourworldindata.org/crop-yields). There are actually five datasets but we are going to focus on global crop yields. 

```{r, out.width="100%", include=FALSE}
library(knitr)
knitr::opts_chunk$set(cache = TRUE, warning = FALSE, message = FALSE, 
                      echo = TRUE, dpi = 300, cache.lazy = FALSE,
                      tidy = "styler", fig.width = 10, fig.height = 6)
options(cli.width = 70, width = 70)

library(tidyverse)
library(tidymodels)
library(scales)
theme_set(theme_light())
```
# Download and view the data
Let's start by downloading the data from the [R4DS Github](https://github.com/rfordatascience/tidytuesday) account and take a look at it. the tidytuesdayR package (by Ellis Hughes, co-host of the screencast 'TidyX') is a handy package for checking out available datasets, exploring the data dictionary in the Rstudio viewer and loading the desired data (and more). Although simple to use, I'm going to stick with the `readr::read_csv()` from the tidyverse to import the data as a tibble (data frame).

```{r}
# Load the data with a readr function
crop_yields <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-09-01/key_crop_yields.csv")

land_use <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-09-01/land_use_vs_yield_change_in_cereal_production.csv") %>% janitor::clean_names()


skimr::skim(crop_yields)

# Here are a few other options for getting to know the data
# glimpse(crop_yields) 
# summary(crop_yields)
# View(crop_yields)
```
# Tidy data: cleaning and reshaping
Let's start with a little data cleaning! `skim()` from `skimr` provides a convenient summary. By looking at the data we can see that there are missing values that need to be removed or imputed but other than that, the data is clean. One thing we should consider is whether the data is in a tidy format where each column is a feature and each row a sample. This long format facilitates fluent analysis with the tidyverse. The function `pivot_longer()` from the tidyr package allows us to reshape the data and prepare it for exploratory data analysis (EDA).   
```{r}
crop_yields %>% count(Year)
# Use janitor package to clean column names and pivot_longer from tidyr to tidy the data.
crop_yield_tidy <- crop_yields %>% 
  janitor::clean_names() %>% 
  rename_all(str_remove, "_tonnes.*") %>% 
  rename(country = entity) %>% 
  pivot_longer(wheat:bananas, names_to = 'crop', values_to = 'yield_hectare') %>% 
  drop_na(yield_hectare)

kable(head(crop_yield_tidy, n = 10))

```
# What can we learn from this dataset?

A great place to start an analysis is by grouping and counting features using the `count()` function. For this project, I have not set out with a specific analysis or machine learning predictive task in mind. By spending time with data, counting and plotting it, we can get a feel for relationships and hopefully generate some insight about what types of questions can we answer. For instance, with this data
```{r}
crop_yield_tidy %>%
  count(year, country, sort = TRUE)

tidy_select <- crop_yield_tidy %>%
  count(country, sort = TRUE) %>%
  filter(n >=500)

crop_yield_tidy %>%
  summarize(max(year)- min(year)) #57 years of recorded data

crop_yield_tidy %>%
  count(year, crop, sort = T)

crop_yield_tidy %>%
  filter(country == 'United States') %>%
  count(crop)
```
# Visual analysis

Now that we have the data in tidy format, let's see what we can learn from visual analysis. We want to look at crop production by country but we can't plot every country. For more refined control of country selection, we could incorporate the code into a Shiny dashboard.

We will use `slice_sample()` can be used to randomly select a subset of the total countries. To get a feel for the data, we will select nine countries. It's worth noting that the randomly selected countries will not always have all crops.
```{r}
# Randomly select 9 country names
# set.seed(1014) if you want to select the same 9 countries each time
crops <- crop_yield_tidy %>% distinct(crop) %>%  pull()
country_random <- crop_yield_tidy %>% 
  # filter(crop %in% crops) %>% 
  select(country) %>% 
  distinct() %>% 
  slice_sample(n = 9) %>% 
  pull()

# another option would be to set the code equal to a set selection of countries: E.g., countries <- c('GUM',	'IND',	'YEM',	'NAM',	'URY', 'USA')

# Plot crop yield per hectare.
library(tidytext)
# plot of USA data only
usa_col_plot <- crop_yield_tidy %>% 
  filter(country == 'United States') %>% 
  mutate(crop = fct_reorder(crop, yield_hectare)) %>%
  ggplot(aes(yield_hectare, crop, fill = crop)) +
  geom_col()  +
  facet_wrap(~country, scales = 'free_y') +
  theme(legend.position =  'none') +
  labs(
    title = 'Crop Yield USA',
    y = 'Crop',
    x = 'tonnes per hectare'
  )
usa_col_plot

# this is what I started with.
# crop_yield_tidy %>% 
#   filter(country == country_random) %>% 
#   mutate(crop = fct_reorder(crop, yield_hectare)) %>%
#   ggplot(aes(yield_hectare, crop, fill = crop)) +
#   geom_col()  +
#   facet_wrap(~country, scales = 'free_y')

column_plot <- crop_yield_tidy %>% 
  filter(country == country_random,
         crop == c('maize', 'potatoes', 'wheat', 'bananas')) %>% 
  mutate(crop = reorder_within(crop, yield_hectare, country)) %>%
  ggplot(aes(yield_hectare, crop, fill = crop)) +
  geom_col()  +
  scale_y_reordered() +
  facet_wrap(~country, scales = 'free') +
  theme(legend.position =  'none') +
  labs(
    title = 'Crop Yield',
    y = 'Crop',
    x = 'yield per hectare'
  )
column_plot
```

```{r}
# Plot crop yield per hectare per year.
line_plot <- crop_yield_tidy %>% 
  filter(year >= 2000,
         country == country_random) %>% 
  mutate(crop = fct_reorder(crop, yield_hectare)) %>% 
  ggplot(aes(year, yield_hectare, color = crop)) +
  geom_line() +
  facet_wrap(~country, scales = 'free_y') +
  labs(
     y = 'yield per hectare',
    title = 'Crop Yield'
  )
line_plot
```


```{r}
# Plot crop yield per hectare per year.
box_plot <- crop_yield_tidy %>% 
  filter(year >= 2000,
         country == country_random) %>% 
  mutate(crop = fct_reorder(crop, yield_hectare)) %>% 
  ggplot(aes(yield_hectare, crop, color = crop)) +
  geom_boxplot() +
  theme(legend.position =  'none') +
  facet_wrap(~country, scales = 'free') +
  labs(
    x = 'Yield',
    title = 'Crop Yield',
    y = ''
  )
box_plot

```

