---
title: Hello Tidyverse and generating many models
author: Gabe Mednick
date: '2020-09-03'
slug: hello-tidyverse
categories: []
tags: []
subtitle: 'Work in progress'
summary: 'This post explores crop yield data with visualization and the lm() function to generate many regression models (a model for each country (25) and crop (4).'
authors: []
lastmod: '2020-10-20'
featured: no
disable_jquery: no
image:
  caption: '[Photo by Jesse Gardner on Unsplash](https://unsplash.com/photos/JYU14tnbfOA)'
  focal_point: ''
  preview_only: no
projects: []
---
# Intro 
This post uses data from the <i class='fab fa-r-project'  style='color: #367588'></i> for Data Science online learning community. Each Tuesday, a dataset is posted and people share their analyses and visualizations on Twitter with the hashtag, #TidyTuesday. This week's dataset (September 1st, 2020) is from [Our World in Data](https://ourworldindata.org/crop-yields). There are actually five datasets but we are going to focus on global crop yields. 

```{r, out.width="100%", include=FALSE}
library(knitr)
knitr::opts_chunk$set(cache = TRUE, warning = FALSE, message = FALSE, 
                      echo = TRUE, dpi = 300, cache.lazy = FALSE,
                      tidy = "styler", fig.width = 10, fig.height = 6)
options(cli.width = 70, width = 70)

library(tidyverse)
library(tidymodels)
library(scales)
library(lubridate)
library(tidytext)
theme_set(theme_light())
```
# Download and view the data
Let's start by downloading the data from the [R4DS Github](https://github.com/rfordatascience/tidytuesday) account and take a look at it. the tidytuesdayR package (by Ellis Hughes, co-host of the screencast 'TidyX') is a handy package for checking out available datasets, exploring the data dictionary in the Rstudio viewer and loading the desired data (and more). Although simple to use, I'm going to stick with the `readr::read_csv()` from the tidyverse to import the data as a tibble (data frame).

```{r}
# Load the data with a readr function
crop_yields <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-09-01/key_crop_yields.csv") %>% 
  janitor::clean_names() 

land_use <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-09-01/land_use_vs_yield_change_in_cereal_production.csv") %>% 
  janitor::clean_names() %>% 
  select(entity, code, year, population = total_population_gapminder) %>% 
  mutate(year = as.numeric(year))

crop_yield <- crop_yields %>% 
  left_join(land_use, by = c('code',  'year', 'entity'))

skimr::skim(crop_yields)

# Here are a few other options for getting to know the data
# glimpse(crop_yields) 
# summary(crop_yields)
# View(crop_yields)
```
# Tidy data: cleaning and reshaping
Let's start with a little data cleaning! `skim()` from `skimr` provides a convenient summary. By looking at the data we can see that there are missing values that need to be removed or imputed but other than that, the data is clean. One thing we should consider is whether the data is in a tidy format where each column is a feature and each row a sample. This long format facilitates fluent analysis with the tidyverse. The function `pivot_longer()` from the tidyr package allows us to reshape the data and prepare it for exploratory data analysis (EDA).   
```{r}
# Use pivot_longer from tidyr to tidy the data.
crop_yield_tidy <- crop_yield %>% 
  rename_all(str_remove, "_tonnes.*") %>% 
  rename(country = entity) %>% 
  pivot_longer(wheat:bananas, names_to = 'crop', values_to = 'yield_hectare') %>% 
  drop_na(yield_hectare) 

# Create a vector of the 25 countries with the largest population
top_pops <- land_use  %>%
  filter(!is.na(code), entity != "World") %>% 
  group_by(entity) %>% 
  filter(year == max(year)) %>% 
  ungroup() %>% 
  slice_max(population, n = 25) %>% 
  pull(entity)


kable(head(crop_yield_tidy, n = 10))

```
# What can we learn from this dataset?

A great place to start any analysis is by grouping and counting predictive features using the `count()` function. It often makes sense to combine `count()` with a bar plot to visualize meaningful distributions. EDA is arguably the most important step in a machine learning workflow because it gives us insight that we can then use to generate a predictive model. When we run a machine learning model, we should already have an intutive sense about what to expect from our model. 

```{r}
crop_yield_tidy %>%
  add_count(year, country, sort = T) %>% 
  filter(country %in% c("Albania", "Africa", "United States")) # Africa has 11 crops, US has 9 and Albania 6. 

crop_yield_tidy %>%
  count(country, sort = TRUE) %>%
  filter(n >=500) # this gives a list of the countries with the most years + crops 11 (crops) * 51 (year) = 561

crop_yield_tidy %>%
  summarize(max(year)- min(year)) #57 years of recorded data

crop_yield_tidy %>% # 249 unique countries
  distinct(country) %>% 
  nrow()

```
# Visual analysis

Now the data is tidy, we are ready to take a closer look at it. Our initial EDA should have given us some ideas about what relationships we might want to pursue with visual analysis. Let's start by comparing crop yield by country. There are too many countries to add to a single plot, so we might need to plot a subset of the data. For more refined control of country selection, we c code into a Shiny dashboard.

We will use `slice_sample()` to randomly select a subset of the total countries. Unless a seed is set, the selection will be different each time we run the code chunk. Let's choose 9 countries to get a feel for crop yields over time without overwhelming our senses. 
```{r}
# Randomly select 9 country names
# set.seed(1014) if you want to select the same 9 countries each time
crops <- crop_yield_tidy %>% distinct(crop) %>%  pull()
country_random <- crop_yield_tidy %>% 
  select(country) %>% 
  distinct() %>% 
  slice_sample(n = 9) %>% 
  pull()

crop_yield_tidy %>% 
  filter(country == 'United States') %>% 
  ggplot(aes(year, yield_hectare, color = crop)) +
  geom_line() +
  facet_wrap(~country, 'free_x') +
  theme(plot.title.position = 'plot') +
  labs(
    title = 'Crop Yield USA',
    y = 'Year',
    x = 'tonnes per hectare per year'
  )


# plot of USA data only
crop_yield_tidy %>% 
  filter(country == 'United States') %>% 
  mutate(crop = fct_reorder(crop, yield_hectare)) %>%
  ggplot(aes(yield_hectare, crop, fill = crop)) +
  geom_col()  +
  facet_wrap(~country, scales = 'free_y') +
  theme(legend.position =  'none',
        plot.title.position = 'plot') +
  labs(
    title = 'Crop Yield USA',
    y = 'Crop',
    x = 'tonnes per hectare'
  )

# Plot crop yield per hectare per year.
crop_yield_tidy %>% 
  filter(country %in% country_random,
         crop %in% c('maize', 'potatoes', 'wheat', 'bananas')) %>% 
  mutate(crop = fct_reorder(crop, yield_hectare)) %>% 
  ggplot(aes(year, yield_hectare, color = crop)) +
  geom_line(size = 1, alpha = 0.5) +
  facet_wrap(~country, scales = 'free_y') +
  theme(plot.title.position = 'plot') +
  labs(
     y = 'yield per hectare',
    title = 'Crop Yield'
  )

```

```{r}
crop_yield_tidy %>% 
  filter(country %in% top_pops,
         crop == c('maize', 'potatoes', 'wheat', 'bananas')) %>% 
  mutate(crop = reorder_within(crop, yield_hectare, country)) %>%
  ggplot(aes(yield_hectare, crop, fill = crop)) +
  geom_col()  +
  scale_y_reordered() +
  facet_wrap(~country, scales = 'free') +
  theme(legend.position = 'none',
        plot.title.position = 'plot') +
  labs(
    title = 'Crop Yield',
    y = 'Crop',
    x = 'yield per hectare'
  )
```
```{r}
crop_yield_tidy %>% 
  filter(country %in% top_pops,
         crop %in% c('maize', 'potatoes', 'wheat', 'bananas')) %>% 
  mutate(crop = fct_reorder(crop, yield_hectare)) %>% 
  ggplot(aes(year, yield_hectare, color = crop)) +
  geom_line(size = 1, alpha = 0.5) +
  theme(plot.title.position = 'plot') +
  facet_wrap(~country, scales = 'free_y') +
  labs(
     y = 'yield per hectare',
    title = 'Crop Yield'
  )

```

Next, let's make a regression model for each of the top 20 countries.
```{r}
library(tidymodels)
library(ggrepel)
crop_yield_tidy %>%
  drop_na() %>%
  filter(country %in% top_pops,
         crop %in% c('maize', 'potatoes', 'wheat', 'bananas')) %>%
  nest(mod_data = c(year, yield_hectare, population)) %>%
  mutate(model = map(mod_data, ~lm(yield_hectare ~ population + year, data = .))) %>%
  mutate(coefs = map(model, tidy)) %>%
  unnest(coefs) %>%
  filter(term == "year") %>%
  mutate(p.value = p.adjust(p.value)) %>% 
  ggplot(aes(estimate, p.value, label = country)) +
  geom_vline(xintercept = 0, lty = 2, 
               size = 1.5, alpha = 0.7, color = "gray50") +
  geom_point(aes(color = crop), alpha = 0.8, size = 2.5, show.legend = FALSE) +
  scale_y_log10() +
  facet_wrap(~crop) +
  geom_text_repel(size = 2.5, family = "Times New Roman") 
  
```


