---
title: "Hourly percentage change for top 100 cryptos"
author: "Gabe Mednick"
date: '2021-06-27'
output:
  html_document:
    df_print: paged
categories: []
tags: []
subtitle: ''
summary: Webscraping to generate custom data for crypto analysis
authors: []
lastmod: ''
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
slug: crypto
---
In this post, I use the httr package to access bitcoin's 24hr price chart from the [CoinGecko](https://www.coingecko.com/en) crypto market [api](https://www.coingecko.com/en/api). The rvest package is used to scrape the hourly percentage change for the top 100 cryptos (by market cap) from CG's website. The cron task scheduler is then set to run the webscraping script every 15 minutes and write the data to a spread sheet, giving me the percentage change on a 15 minute interval.

Once we have collected enough data, I plan to use it for predicting the percentage change for Bitcoin. In order for this to work, the task scheduler and webscraping script will need to run in the cloud. Ideally, this post will be turned into a Shiny app so that it can be used for real time analysis and price prediction.

Currently this post shows several ways to view the hourly percentage changes as a static snapshot. I've categorized the hourly percentage change as positive and negative, as well as creating a more nuanced scale with the following categories: `ice age (< -6%), freezing (-4-6%), cold (-2-4%), chilly (-0-2%), neutral (0%), luke warm (0-2%), warm (2-4%), toasty (4-6%), hot (>6%)`.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
```

```{r}
library(tidyverse)
library(tidyr)
library(plotly)
library(lubridate)
library(httr)
library(jsonlite)
library(viridis)
library(scales)
library(rvest)
library(cronR)
library(patchwork)
theme_set(theme_light())
```

```{r}
url_cg1 <- "https://www.coingecko.com/en"
url_cg2 <- "https://www.coingecko.com/en?page=2"

names <- read_html(url_cg1) %>% 
  html_nodes('.font-bold.tw-justify-between') %>% 
  html_text() %>% 
  as_tibble()

names2 <- read_html(url_cg2) %>% 
  html_nodes('.font-bold.tw-justify-between') %>% 
  html_text() %>% 
  as_tibble()

hourly <- read_html(url_cg1) %>% 
  html_nodes('.change1h span') %>% 
  html_text() %>% 
  as_tibble() 
 
hourly2 <- read_html(url_cg2) %>% 
  html_nodes('.change1h span') %>% 
  html_text() %>% 
  as_tibble()

hourly_df <- bind_cols(names, hourly) %>% 
  rename("coin" = `value...1`, "hourly" = `value...2`) %>% 
  mutate(coin = str_remove_all(coin, "\n"),
         hourly = parse_number(hourly))

# hourly_df2 <- bind_cols(names2, hourly2) %>% 
#   rename("coin" = `value...1`, "hourly" = `value...2`) %>% 
#   mutate(coin = str_remove_all(coin, "\n"),
#          hourly = parse_number(hourly))
# 
# hourly_df200 <- bind_rows(hourly_df, hourly_df2)

granular_hourly <- hourly_df %>% 
  mutate(change = case_when(hourly > 0 ~ "positive",
                            hourly <= 0 ~ "negative"),
         granular = case_when((hourly > 6) ~ "hot (>6%)",
                              (hourly > 4) & (hourly <= 6) ~ "toasty (4-6%)",
                              (hourly > 2) & (hourly <= 4) ~ "warm (2-4%)",
                              (hourly > 0) & (hourly <= 2) ~ "luke warm (0-2%)",
                              (hourly == 0) ~ "neutral (0%)",
                              (hourly <= 0) & (hourly >= -2)  ~ "chilly (-0-2%)",
                              (hourly <= -2) & (hourly > -4) ~ "cold (-2-4%)",
                              (hourly <= -4) &(hourly > -6) ~ "freezing (-4-6%)",
                              TRUE ~ "ice age (< -6%)"),
         granular = factor(granular, 
                           levels = c("ice age (< -6%)", "freezing (-4-6%)", "cold (-2-4%)", 
                                      "chilly (-0-2%)", "neutral (0%)", "luke warm (0-2%)", "warm (2-4%)", 
                                      "toasty (4-6%)", "hot (>6%)")),
         time_stamp =  as_datetime(Sys.time()))


# write_tsv(
#   granular_hourly,
#   'granular_15min.tsv',
#   na = "NA",
#   col_names = TRUE,
#   append = TRUE
# )
library(viridis)
granular_hourly %>% 
  mutate(granular = fct_rev(granular)) %>% 
  group_by(change, granular) %>% 
  count(change) %>% 
  ggplot(aes(change, n, fill = granular)) +
  geom_col() +
  #scale_fill_viridis_d(option = "viridis") +
  #scale_fill_discrete(guide = guide_legend(reverse = TRUE)) +
  labs(title =  paste0("Hourly change for top 100 cryptos: ", format(Sys.time(), "%B %d, %Y @ %I:%M %p", tz = "PST8PDT")),
       fill = "hourly % change",
       y = "", x = "") +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank())

#ggsave("hrly-barchart.png")
```

```{r}
library(hpackedbubble)

hpackedbubble(granular_hourly$granular, 
              granular_hourly$coin, granular_hourly$hourly,
              title = "Hourly percentage change",
              titleColor = "#F0FFF0",
              pointFormat = "<br>{point.name} </br> Volume: {point.value}",
              dataLabelsFilter = 100,
              packedbubbleMinSize = "50%",
              packedbubbleMaxSize = "250%",
              theme = "darkunica",
              packedbubbleZMin = 0,
              packedbubbleZmax = 10000, split = 0,
              gravitational = 0.02,
              parentNodeLimit = 1,
              dragBetweenSeries = 0,
              width = "100%",
              dataLabelsColor = "white")

```


```{r}
pct_change_plot <- granular_hourly %>% 
  count(granular) %>% 
  ggplot(aes(granular, n, fill = granular)) +
  geom_col() +
  labs(title = paste0("Hourly percent change for top 100 cryptos: ", format(Sys.time(), "%B %d @ %I:%M %p", tz = "PST8PDT")),
       x = "", y = "Coin count") + 
  theme(legend.position = "none")

pct_change_plot
```


```{r}
pct_change_btc_eth_removed <- granular_hourly %>% 
  filter(!coin %in% c("Bitcoin", "Ethereum")) %>% 
  count(granular) %>% 
  ggplot(aes(granular, n, fill = granular)) +
  geom_col() +
  labs(title = "Hourly percent change for top 98 cryptos",
       subtitle = "Bitcoin and Ethereum removed",
       x = "", y = "Coin count") + 
  theme(legend.position = "none") 

pct_change_btc_eth_removed

#ggsave("pct_change_plot.png")
```


```{r}
api_ohlc <- "https://api.coingecko.com/api/v3/coins/bitcoin/ohlc?vs_currency=usd&days=30"

ohlc <- GET(api_ohlc)
ohlc <- content(ohlc, as = 'text')

ohlc_df <- fromJSON(ohlc, flatten = TRUE) %>% 
  as_tibble()

names(ohlc_df) <- c('time', 'open', 'high', 'low', 'close')


ohlc_df %>% 
  mutate(time = time/1000,
         time = as_datetime(time)) %>% 
  select(time, open) %>% 
  arrange(time) %>% 
  ggplot(aes(time, open, color = open)) +
  geom_line() +
  theme(legend.position = "none") +
  labs(title = "Bitcoin monthly price chart",
       x = "", y = "Value (USD)")

```

```{r}
btc_url = "https://api.coingecko.com/api/v3/coins/bitcoin/market_chart?vs_currency=usd&days=1"

get_historical <- function(url) {
    url %>% 
    GET() %>% 
    content(., as = "text") %>% 
    fromJSON(., flatten = TRUE)  
}

library(hms)
#test function
btc_chart <- get_historical(btc_url) %>% 
  do.call(data.frame, .) %>%
          as_tibble() %>% 
  select('date' = prices.1, 'price' = prices.2,
         'market_cap' = market_caps.2, 'volume' = total_volumes.2) %>% 
  mutate(date = date/1000,
         date = as.POSIXct(date, origin="1970-01-01"),
         time = format(date, format = '%H:%M:%S', tz = "PST8PDT") %>% as_hms())

btc_chart %>% 
  ggplot(aes(date, price, color = volume, group = 1)) +
  geom_line() +
  labs(title = "24hr Bitcoin price chart") +
  scale_x_datetime(
    name = "date",
    labels = label_date("%I:%M %p", tz = "PST8PDT"))

```

```{r}
q_minute <- btc_chart %>% 
  filter(date >= '2021-06-26 14:54:57') 

btc_minute_plot <- q_minute %>% 
  ggplot(aes(date, price, color = volume, group = 1)) +
  geom_line() +
  labs(title = paste0("Bitcoin price chart for ", format(Sys.time(), "%B %d %Y")), x = "", y = "") +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()) +
  scale_x_datetime(
    name = "",
    labels = label_date("%I:%M %p", tz = "PST8PDT")) +
  scale_y_continuous(labels = scales::dollar) 

#ggsave("btc-hrly-pct.png", btc_minute_plot, bg = "transparent")

granular_15min <- read_tsv("granular_15min.tsv") %>% 
  mutate(time_stamp = str_remove(time_stamp, "Z"),
    time_stamp = ymd_hms(time_stamp)) %>% 
  filter(coin != "coin") 

bar_plot <- granular_15min  %>% 
  group_by(time_stamp, granular, change) %>% 
  count(change) %>% 
  ggplot(aes(time_stamp, n, fill = granular)) +
  geom_col() +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()) +
  scale_x_datetime(
    name = "time",
    labels = label_date("%I:%M %p", tz = "PST8PDT")) +
  labs(y = NULL)

#ggsave("alt100-15min-pct.png", bar_plot, bg = "transparent")

btc_minute_plot / bar_plot
```

